---
title: "Prediction of closest cluster for measured features"
author: "Inga Haalck, supervised by Carolin Huber"
date: "2024-07-24"
output: html_document
---

# Loading of required of packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(tibble)
library(janitor)
library(ggplot2)
```

# Data that needs to be loaded

```{r}
mean_traj <- read.csv("05_a_kml_pred.csv") # table with centers of clusters
NTS_df <- read.csv("02_c_datawrangling_NTS_readyforkml.csv") # non-target features that should be assigned to a cluster, columns 9:80 contains signal intensity values
```

# further definitions

```{r}
# Define cluster labels
cluster_labels <- c("A", "B", "C", "D", "E")
```

# define predict function
```{r}
################################################################################
#  define predict function
predict_cluster <- function(NTS_subset_norm, mean_traj2, cluster_labels) {
  # make sure NTS_subset_norm and mean_traj2 have the same number of columns
  if (ncol(NTS_subset_norm) != ncol(mean_traj2)) {
    stop("Number of columns in both data sets must match (number of time points).")
  }

  # make sure the length of cluster_labels matches the number of clusters
  if (length(cluster_labels) != nrow(mean_traj2)) {
    stop("Number of cluster labels must match the number of clusters.")
  }

  # make a new matrix to hold the calculated distances (number of features as rows, number of clusters as columns)
  num_data_points <- nrow(NTS_subset_norm)
  num_clusters <- nrow(mean_traj2)
  distances <- matrix(NA, nrow = num_data_points, ncol = num_clusters)

  # iterate over each data point (i) and cluster (j) and calculate the squared euclidean distance between each point and       cluster & store the distance in matrix 'distances'
  for (i in 1:num_data_points) {
    for (j in 1:num_clusters) {
      distances[i, j] <- sum((NTS_subset_norm[i, ] - mean_traj2[j, ])^2)
    }
  }

  # determine the closest cluster for each data point
  closest_cluster_indices <- apply(distances, 1, which.min)

  # give the closest cluster label to each features
  closest_clusters <- cluster_labels[closest_cluster_indices]

  # create a data frame with distances and the closest cluster assignments
  distances_df <- as.data.frame(distances)
  distances_df$ClosestCluster <- closest_clusters

  return(distances_df)
}
###############################################################################
# perform calculation (only run once - takes time)
# distances_df1 <- predict_cluster(NTS_subset_norm, mean_traj2, cluster_labels)
# Save the data frame as an R data file for later use
# save(distances_df1, file = "distances_df_new6_5_morefeatures_sqrt.RData")
################################################################################
```

# Data Wrangling for NTS data 

## 1. Remove duplicated IDs and remove columns that are not needed 

```{r}
# make new column with featureID_ESImode to use later as unique identification of features
NTS_df <- NTS_df %>%
  mutate(id_ESI = paste(id, ESI_mode, sep = "_"), .before = ESI_mode)

# Step 1: Calculate the mean of columns 9 to 80 for each row
NTS_df_mean <- NTS_df %>%
  rowwise() %>%
  mutate(mean_value = mean(c_across(9:80), na.rm = TRUE)) %>%
  ungroup()

# Step 2: Group by 'id' and keep the row with the highest mean value within each group
NTS_df_dup <- NTS_df_mean %>%
  group_by(id) %>%
  filter(mean_value == max(mean_value)) %>%
  ungroup() %>%
  select(-mean_value)

# Step 3: Remove unnecessary information
NTS_df2 <- NTS_df_dup %>%
  select(-id, -ESI_mode, -rt, -mz, -target, -speclib, -urine)

feat_ids <- NTS_df2$id_ESI

# change colnames to match mean_traj data frame
colnames(NTS_df2) <- colnames(mean_traj)
```

## 2. Replace NAs by 1/4 of min value (same as for targets)

```{r}
get_min <- apply(NTS_df2[, 2:ncol(NTS_df2)], 1, function(x) min(x, na.rm = TRUE))

for (i in 1:nrow(NTS_df2)) {
  xx <- which(is.na(NTS_df2[i, ]))
  NTS_df2[i, xx] <- get_min[i] / 4
}
```

## 3. Scale & log transformation (same as for target df used for kml) 

```{R}
NTS_subset <- NTS_df2 %>%
  select(-X)

NTS_subset_norm <- t(scale(t(log2(NTS_subset))))
NTS_subset_norm <- as.data.frame(NTS_subset_norm)
```

--------------------------------------------------------------------------------

# Cluster Prediction 

## 1. Calculate Euclidean distance between clusters and features and choose the closest cluster 

```{r}
# remove grouping variable (cluster names) out of data frame
mean_traj2 <- mean_traj %>%
  select(-X)

# Load the data frame back into R
load("distances_df.RData")

# Squared Euclidean distances
distances_df2 <- distances_df %>%
  mutate(ids = feat_ids)
```

## 2. Filter results (based on distance ratio or distance threshold)

```{r}
# Remove the last two columns
distances_df_clustersonly <- distances_df2 %>%
  select(1:(ncol(distances_df2) - 2))

# Calculate the minimum distance for each row
min_distances_2 <- distances_df_clustersonly %>%
  rowwise() %>%
  mutate(MinDistance = min(c_across(everything()), na.rm = TRUE)) %>%
  pull(MinDistance)

# Calculate the second minimum distance for each row
second_min_distances_2 <- distances_df_clustersonly %>%
  rowwise() %>%
  mutate(SecondMinDistance = {
    sorted_row <- sort(c_across(everything()))
    if (length(sorted_row) > 1) sorted_row[2] else NA_real_
  }) %>%
  pull(SecondMinDistance)

# Add closest and 2nd closest distances to distances_df dataframe
distances_df2 <- distances_df2 %>%
  mutate(
    MinDistance = min_distances_2,
    SecondMinDistance = second_min_distances_2
  )

# Set columns to numeric to enable calculations
distances_df2 <- distances_df2 %>%
  mutate(MinDistance = as.numeric(MinDistance)) %>%
  mutate(SecondMinDistance = as.numeric(SecondMinDistance)) %>%
  mutate(ratio = SecondMinDistance / MinDistance)

# Apply the filtering criterion: ratio > 1.5
filtered_distances_df2 <- distances_df2 %>%
  filter(ratio > 1.5) %>%
  filter(MinDistance < 44)

# Save filtered feature ids
filtered_feat <- filtered_distances_df2$ids

# Save closest clusters for each feature
closest_cluster <- filtered_distances_df2$ClosestCluster
```

# export ids, rt & mz for SIRIUS

```{r}
# all features
NTS_df_short <- NTS_df_dup %>%
  select(id_ESI, rt, mz) %>%
  rename(ids = id_ESI)

distances_short <- distances_df2 %>%
  select(ClosestCluster, ids)

NTS_clustered <- NTS_df_short %>%
  inner_join(distances_short, by = "ids")

NTS_clustered <- NTS_clustered %>%
  separate(ids, into = c("id", "ESI"), sep = "_")
```

# export filtered features 

```{r}
NTS_df_short_filtered <- NTS_df_dup %>%
  filter(id_ESI %in% filtered_feat) %>%
  select(id_ESI, rt, mz) %>%
  rename(ids = id_ESI)

distances_short_filtered <- distances_df2 %>%
  filter(ids %in% filtered_feat) %>%
  select(ClosestCluster, ids)

NTS_clustered_filtered <- NTS_df_short_filtered %>%
  inner_join(distances_short_filtered, by = "ids")

NTS_clustered_filtered <- NTS_clustered_filtered %>%
  separate(ids, into = c("id", "ESI"), sep = "_")
```
