---
title: "Cluster analysis Markkleeberg pos on 3 days"
author: "Carolin Huber & Inga Haalck"
date: "2023-07-29"
output: html_document
---

# Installation of packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(plotly)
library(scales)
library(reshape2)
library(data.table)
library(factoextra)
library(cluster)
library(stats)
library(kml)
library(knitr)
```

# Data that needs to be loaded

```{R}
dat <- read.csv("02_b_datawrangling_targets_readyforkml.csv") # load annotated table
```

# Define filenames for output storage

```{R}
cluster_rds <- "cluster_model.rds" # storage of the calculated cluster, can be used later to safe calculation time
trajectories_cluster <- "05_a_kml_pred_test.csv" # cluster centers used for further prediction script
cluster_annotation_csv <- "05_a_kml_allbatches_clusterannotations_test.csv" # annotation separated by clusters
cluster_assignment <- "05_a_kml_allbatches_test.csv" # file with cluster for each feature
```

# Data cleanup

Replace NA by 1/4 of min value
```{r}
get_min <- apply(dat, 1, function(x) min(x, na.rm = T))

get_min <- apply(dat[, 4:ncol(dat)], 1, function(x) min(x, na.rm = TRUE))

for (i in 1:nrow(dat)) {
  xx <- which(is.na(dat[i, ]))
  dat[i, xx] <- get_min[i] / 4
}
```

Subset the dataframe to exclude the first two columns which contain metadata and not SI values
```{r}
compounds_ids <- dat[, 1]
compounds_names <- dat[, 2]
compound_class <- dat[, 3]
data_subset <- dat[, -(1:4)]
```

# Perform longitudinal k-means clustering (kml) for 72 hours 
```{R}
data_subset_norm <- t(scale(t(log2(data_subset))))

# calculate, run separately and added to markdown as rds file
set.seed(1)
event_long <- clusterLongData(data_subset_norm, idAll = c(1:nrow(data_subset_norm)), time = c(1:ncol(data_subset_norm)), varNames = "Signal abundance")

# perform kml analysis
kml(event_long,
  nbClusters = 2:6,
  nbRedrawing = 10,
  toPlot = "none",
  parAlgo = parALGO(saveFreq = 100, distanceName = "euclidean", distance = diss.CORT)
)

# save kml results
saveRDS(event_long, file = cluster_rds)
```

# optional: re-load cluster results  
```{r}
# Load the R object from the file
event_long <- readRDS(cluster_rds)
```

```{r}
# plot all criteria to choose cluster number
plotAllCriterion(event_long)
# chose cluster number
no_clusters <- 5
# plot overview results for selected number of clusters
plot(event_long, no_clusters, parMean = parMEAN(cex = 0.3), parTraj = parTRAJ(col = "clusters"), toPlot = "traj")
```

```{r}
# extract cluster members
dat$cluster <- getClusters(event_long, no_clusters)
# get probabilities for cluster fitting/ cluster evaluation
dat$clustProp <- getBestPostProba(event_long, nbCluster = no_clusters, clusterRank = 1)
hist(dat$clustProp)
# filter by cluster probability
dat <- dat[which(dat$clustProp > 0.9), ]
```

# extract cluster centroids 
```{r}
mean.trajectories <- calculTrajMean(event_long["traj"], event_long["c5"][[1]]["clusters"])

mean_traj <- as.data.frame(mean.trajectories)

mean_traj_long <- mean_traj %>%
  rownames_to_column(var = "cluster")

mean_traj_long2 <- mean_traj_long %>%
  pivot_longer(cols = starts_with("t"), names_to = "hour", values_to = "value") %>%
  mutate(hour = as.numeric(stringr::str_remove(hour, "t")))

ggplot(mean_traj_long2, aes(x = hour, y = value, color = cluster)) +
  geom_line() +
  labs(x = "Hour", y = "Value", color = "Cluster") +
  scale_y_continuous(limits = c(-6, 6)) +
  theme_minimal()

# export
write.csv(mean_traj, trajectories_cluster)
```

# Extraction of cluster annotations 
```{R} 
# Initialize an empty list to store annotations for each cluster
cluster_annotations <- list()

# Iterate over each cluster
for (i in 1:no_clusters) {
  # Subset the data for the current cluster
  x <- dat[which(dat$cluster == LETTERS[i]), ]
  x <- x[grep("[a-z]", x$compounds_names), ]
  # Add the annotations to the list
  cluster_annotations[[i]] <- data.frame(Cluster = LETTERS[i], compounds_names = x$compounds_names)
}

# Combine the annotations from all clusters into a single data frame
annotations_df <- do.call(rbind, cluster_annotations)

# Print the first few rows of the new data frame
head(annotations_df)
for (i in 1:no_clusters) {
  # get annotations summarized
  x <- dat[which(dat$cluster == LETTERS[i]), ]
  x <- x[grep("[a-z]", x$compounds_names), ]
  # summarize output
  message("\n ************************** Target annotations in cluster ", LETTERS[i], " **************************")
  print(kable(x$compounds_names))
}

annotations_names <- annotations_df$compounds_names

kml_compounds <- dat %>%
  filter(compounds_names %in% annotations_names) %>%
  select(compounds_names, compound.class.level.1) %>%
  rename(compounds_class = compound.class.level.1)

# combine the two data frames
kml_compoundclasses <- annotations_df %>%
  left_join(kml_compounds, by = "compounds_names")

# filter out only compounds that remain from the whole standard list & make new data frame with additional information
filtered_classes <- standards %>%
  filter(Standard %in% annotations_names) %>%
  select(Standard, compound.class.level.1, compound.class.level.2, compound.class.level.3) %>%
  rename(compounds_names = Standard)

# combine the two data frames
kml_compoundclasses <- kml_compoundclasses %>%
  left_join(filtered_classes, by = "compounds_names")

fwrite(kml_compoundclasses, file = cluster_annotation_csv)
``` 

# Extract Cluster Assignment 
```{r}
kml_compounds_df <- dat %>%
  filter(compounds_names %in% annotations_names) %>%
  left_join(kml_compounds, by = "compounds_names") %>%
  select(compounds_names, compounds_class, cluster, everything()) %>%
  select(-clustProp) %>%
  select(-compound.class.level.1)

fwrite(kml_compounds_df, file = cluster_assingment)
```
